---
title: "Padova 2020"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---

## Setup

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE, 
  dev = "png",
  dpi = 150,
  fig.align = "center",
  comment = NA
)
library(rstan)
library(dplyr)
library(lubridate)
library(ggplot2)
library(bayesplot)

theme_set(bayesplot::theme_default())

# seed for R's pseudo-RNGs, not Stan's
set.seed(1123) 
```

## The problem

### Background

Imagine that you are a statistician or data scientist working as an independent
contractor. One of your clients is a company that owns many residential buildings 
throughout New York City. The property manager explains that they are concerned about the number
of cockroach complaints that they receive from their buildings. Previously
the company has offered monthly visits from a pest inspector as a solution to
this problem. While this is the default solution of many property managers in
NYC, the tenants are rarely home when the inspector visits, and so the manager
reasons that this is a relatively expensive solution that is currently not very
effective.

One alternative to this problem is to deploy long term bait stations. In this
alternative, child and pet safe bait stations are installed throughout the
apartment building. Cockroaches obtain quick acting poison from these stations
and distribute it throughout the colony. The manufacturer of these bait stations
provides some indication of the space-to-bait efficacy, but the manager suspects
that this guidance was not calculated with NYC roaches in mind. NYC roaches, the
manager rationalizes, have more hustle than traditional roaches; and NYC
buildings are built differently than other common residential buildings in the
US. This is particularly important as the unit cost for each bait station per
year is quite high.

### The goal

A subset of the company's buildings have been randomly selected for an experiment: 

* At the beginning of each month, a pest inspector randomly places a number of
bait stations throughout the building, without knowledge of the current
cockroach levels in the building
* At the end of the month, the manager records
the total number of cockroach complaints in that building. 

We will model the number of complaints as a function
of the number of traps and other predictors. We want to find a good model on the ground of some posterior predictive diagnostics.

## The data

The data provided to us is in a file called `pest_data.RDS`. Let's
load the data and see what the structure is:

```{r load-data}
pest_data <- readRDS('data/pest_data.RDS')
str(pest_data)
```

We have access to the following fields: 

* `complaints`: Number of complaints per building per month
* `building_id`: The unique building identifier
* `traps`: The number of traps used per month per building
* `date`: The date at which the number of complaints are recorded
* `live_in_super`: An indicator for whether the building as a live-in super
* `age_of_building`: The age of the building
* `total_sq_foot`: The total square footage of the building
* `average_tenant_age`: The average age of the tenants per building
* `monthly_average_rent`: The average monthly rent per building
* `floors`: The number of floors per building

First, let's see how many buildings we have data for:

```{r describe-data}
N_buildings <- length(unique(pest_data$building_id))
N_buildings
```

And make some plots of the raw data: 

```{r data-plots}
ggplot(pest_data, aes(x = complaints)) + 
  geom_bar()

ggplot(pest_data, aes(x = traps, y = complaints, color = live_in_super == TRUE)) + 
  geom_jitter()
```

```{r, data-plots-ts, fig.width = 6, fig.height = 8}
ggplot(pest_data, aes(x = date, y = complaints, color = live_in_super == TRUE)) + 
  geom_line(aes(linetype = "Number of complaints")) + 
  geom_point(color = "black") + 
  geom_line(aes(y = traps, linetype = "Number of traps"), color = "black", size = 0.25) + 
  facet_wrap(~building_id, scales = "free", ncol = 2, labeller = label_both) + 
  scale_x_date(name = "Month", date_labels = "%b") + 
  scale_y_continuous(name = "", limits = range(pest_data$complaints)) + 
  scale_linetype_discrete(name = "") + 
  scale_color_discrete(name = "Live-in super")
```


The first question we'll look at is just whether the number of complaints per
building per month is associated with the number of bait stations per building
per month, ignoring the temporal and across-building variation (we'll come back
to those sources of variation later in the document). That requires only two
variables, $\textrm{complaints}$ and $\textrm{traps}$. How should we model the
number of complaints?


## Modeling count data : Poisson distribution

We already know some rudimentary information about what we should expect. The
number of complaints over a month should be either zero or an integer. The
property manager tells us that it is possible but unlikely that number of
complaints in a given month is zero. Occasionally there are a very large number
of complaints in a single month. A common way of modeling this sort of skewed,
single bounded count data is as a Poisson random variable. One concern about
modeling the outcome variable as Poisson is that the data may be
over-dispersed, but we'll start with the Poisson model and then check 
whether over-dispersion is a problem by comparing our model's predictions 
to the data.

### Model 

Given that we have chosen a Poisson regression, we define the likelihood to be
the Poisson probability mass function over the number bait stations placed in
the building, denoted below as `traps`. This model assumes that the mean and
variance of the outcome variable `complaints` (number of complaints) is the
same. We'll investigate whether this is a good assumption after we fit the 
model.

For building $b = 1,\dots,10$ at time (month) $t = 1,\dots,12$, we have

$$
\begin{align*}
\textrm{complaints}_{b,t} & \sim \textrm{Poisson}(\lambda_{b,t}) \\
\lambda_{b,t} & = \exp{(\eta_{b,t})} \\
\eta_{b,t} &= \alpha + \beta \, \textrm{traps}_{b,t}
\end{align*}
$$

Let's encode this probability model in a Stan program. 


Now we can fit the model by calling the `sampling()` function.


### Fit with real data 

To fit the model to the actual observed data we'll first create a list to pass
to Stan using the variables in the `pest_data` data frame:

```{r stan-data}
stan_dat_simple <- list(
  N = nrow(pest_data), 
  complaints = pest_data$complaints,
  traps = pest_data$traps
)
str(stan_dat_simple)
```

As we have already compiled the model, we can jump straight to sampling from it.

```{r fit_P_real_data, cache=TRUE}
comp_model_P <- stan_model('stan_programs/simple_poisson_regression.stan')
fit_P_real_data <- sampling(comp_model_P, data = stan_dat_simple)
```

and printing the parameters. What do these tell us? 

```{r results_simple_P}
print(fit_P_real_data, pars = c('alpha','beta'))
```

We can also plot the posterior distributions: 

```{r hist_simple_P}
mcmc_hist(as.matrix(fit_P_real_data, pars = c('alpha','beta')))
mcmc_scatter(as.matrix(fit_P_real_data, pars = c('alpha','beta')), alpha = 0.2)
```

As we expected, it appears the number of bait stations set in a building is
associated with the number of complaints about cockroaches that were made in the
following month. However, we still need to consider how well the model fits.


### Posterior predictive checking

```{r}
y_rep <- as.matrix(fit_P_real_data, pars = "y_rep")
```

```{r marginal_PPC}
ppc_dens_overlay(y = stan_dat_simple$complaints, y_rep[1:200,])
```

The simulated
datasets is not as dispersed as the observed data and don't seem to capture the
rate of zeros in the observed data. The Poisson model may not be sufficient for
this data.

Let's explore this further by looking directly at the proportion of zeros in the
real data and predicted data.
```{r}
prop_zero <- function(x) mean(x == 0)
ppc_stat(y = stan_dat_simple$complaints, yrep = y_rep, stat = "prop_zero")
```
The plot above shows the observed proportion of zeros (thick vertical line) and
a histogram of the proportion of zeros in each of the simulated data sets. It is
clear that the model does not capture this feature of the data well at all.

This next plot is a plot of the standardised residuals of the observed vs predicted number of complaints. 

```{r}
mean_y_rep <- colMeans(y_rep)
std_resid <- (stan_dat_simple$complaints - mean_y_rep) / sqrt(mean_y_rep)
qplot(mean_y_rep, std_resid) + hline_at(2) + hline_at(-2)
```

As you can see here, it looks as though we have more positive residuals than negative,
which indicates that the model tends to underestimate the number of complaints
that will be received.

We can also view how the predicted number of complaints varies with the number
of traps. From this we can see that the model doesn't seem to fully capture the
data.

```{r}
ppc_intervals(
  y = stan_dat_simple$complaints, 
  yrep = y_rep,
  x = stan_dat_simple$traps
) + 
  labs(x = "Number of traps", y = "Number of complaints")
```

Specifically, the model doesn't capture the tails of the observed data very
well.


## Expanding the model: multiple predictors

Modeling the relationship between complaints and bait stations is the simplest 
model. We can expand the model, however, in a few ways that will be beneficial 
for our client. Moreover, the manager has told us that they expect there are a
number of other reasons that one building might have more roach complaints than
another.

### Interpretability

Currently, our model's mean parameter is a rate of complaints per 30 days, but
we're modeling a process that occurs over an area as well as over time. We have
the square footage of each building, so if we add that information into the
model, we can interpret our parameters as a rate of complaints per square foot
per 30 days.

$$
\begin{align*}
\textrm{complaints}_{b,t} & \sim \textrm{Poisson}(\textrm{sq_foot}_b\,\lambda_{b,t}) \\
\lambda_{b,t} & = \exp{(\eta_{b,t} )} \\
\eta_{b,t} &= \alpha + \beta \, \textrm{traps}_{b,t}
\end{align*}
$$

The term $\text{sq_foot}$ is called an exposure term. If we log the term, we can 
put it in $\eta_{b,t}$:

$$
\begin{align*}
\textrm{complaints}_{b,t} & \sim \textrm{Poisson}(\lambda_{b,t}) \\
\lambda_{b,t} & = \exp{(\eta_{b,t} )} \\
\eta_{b,t} &= \alpha + \beta \, \textrm{traps}_{b,t} + \textrm{log_sq_foot}_b
\end{align*}
$$

A quick test shows us that there appears to be a relationship between the square
footage of the building and the number of complaints received:

```{r}
ggplot(pest_data, aes(x = log(total_sq_foot), y = log1p(complaints))) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```

Using the property manager's intuition, we include two extra pieces of
information we know about the building - the (log of the) square floor space and
whether there is a live in super or not - into both the simulated and real data.
 
```{r}
stan_dat_simple$log_sq_foot <- log(pest_data$total_sq_foot/1e4)
stan_dat_simple$live_in_super <- pest_data$live_in_super
```

### Stan program for Poisson multiple regression

Now we need a new Stan model that uses multiple predictors.


And then compile and fit the model we wrote for the multiple regression.

### Fit the real data

Now let's use the real data and explore the fit.

```{r fit_mult_P_real_dat}
comp_model_P_mult <- stan_model('stan_programs/multiple_poisson_regression.stan')
fit_model_P_mult_real <- sampling(comp_model_P_mult, data = stan_dat_simple)
y_rep <- as.matrix(fit_model_P_mult_real, pars = "y_rep")
ppc_dens_overlay(stan_dat_simple$complaints, y_rep[1:200,])
```

This again looks like we haven't captured the smaller counts very well, nor
have we captured the larger counts.


```{r}
prop_zero <- function(x) mean(x == 0)
ppc_stat(y = stan_dat_simple$complaints, yrep = y_rep, stat = "prop_zero", binwidth = 0.01)
```

We're still severely underestimating the proportion of zeros in the data. Ideally
this vertical line would fall somewhere within the histogram.

We can also plot uncertainty intervals for the predicted complaints for different
numbers of traps.

```{r}
ppc_intervals(
  y = stan_dat_simple$complaints, 
  yrep = y_rep,
  x = stan_dat_simple$traps
) + 
  labs(x = "Number of traps", y = "Number of complaints")
```

We can see that we've increased the tails a bit more at the larger numbers of traps
but we still have some large observed numbers of complaints that the model
would consider extremely unlikely events. 
 
## Modeling count data with the Negative Binomial

When we considered modelling the data using a Poisson, we saw that the model
didn't appear to fit as well to the data as we would like. In particular the
model underpredicted low and high numbers of complaints, and overpredicted the
medium number of complaints. This is one indication of over-dispersion, where
the variance is larger than the mean. A Poisson model doesn't fit over-dispersed
count data very well because the same parameter $\lambda$, controls both the
expected counts and the variance of these counts. The natural alternative to
this is the negative binomial model:

$$
\begin{align*}
\text{complaints}_{b,t} & \sim \text{Neg-Binomial}(\lambda_{b,t}, \phi) \\
\lambda_{b,t} & = \exp{(\eta_{b,t})} \\
\eta_{b,t} &= \alpha + \beta \, {\rm traps}_{b,t} + \beta_{\rm super} \, {\rm super}_{b} + \text{log_sq_foot}_{b}
\end{align*}
$$

In Stan the negative binomial mass function we'll use is called 
$\texttt{neg_binomial_2_log}(\text{ints} \, y, \text{reals} \, \eta, \text{reals} \, \phi)$ 
in Stan. Like the `poisson_log` function, this negative binomial mass function
that is parameterized in terms of its log-mean, $\eta$, but it also has a
precision $\phi$ such that

$$
\mathbb{E}[y] \, = \lambda = \exp(\eta)
$$

$$
\text{Var}[y] = \lambda + \lambda^2/\phi = \exp(\eta) + \exp(\eta)^2 / \phi.
$$ 

As $\phi$ gets larger the term $\lambda^2 / \phi$ approaches zero and so 
the variance of the negative-binomial approaches $\lambda$, i.e., the
negative-binomial gets closer and closer to the Poisson.

### Stan program for negative-binomial regression


### Fit to real data and check the fit

```{r runNB}
comp_model_NB <- stan_model('stan_programs/multiple_NB_regression.stan')
fitted_model_NB <- sampling(comp_model_NB, data = stan_dat_simple)
samps_NB <- rstan::extract(fitted_model_NB)
```

Let's look at our predictions vs. the data.

```{r ppc-full}
y_rep <- samps_NB$y_rep
ppc_dens_overlay(stan_dat_simple$complaints, y_rep[1:200,])
```

It appears that our model now captures both the number of small counts better
as well as the tails. 

Let's check if the negative binomial model does a better job capturing
the number of zeros: 

```{r}
ppc_stat(y = stan_dat_simple$complaints, yrep = y_rep, stat = "prop_zero")
```

These look OK, but let's look at the standardized residual plot.

```{r}
mean_inv_phi <- mean(samps_NB$inv_phi)
mean_y_rep <- colMeans(y_rep)
std_resid <- (stan_dat_simple$complaints - mean_y_rep) / sqrt(mean_y_rep + mean_y_rep^2*mean_inv_phi)
qplot(mean_y_rep, std_resid) + hline_at(2) + hline_at(-2)
```

Looks OK, but we still have some very large _standardized_ residuals. This might be
because we are currently ignoring that the data are clustered by buildings, and
that the probability of roach issue may vary substantially across buildings.


Check predictions by number of traps: 

```{r}
ppc_intervals(
  y = stan_dat_simple$complaints, 
  yrep = y_rep,
  x = stan_dat_simple$traps
) + 
  labs(x = "Number of traps", y = "Number of complaints")
```

We haven't used the fact that the data are clustered by building yet. A posterior 
predictive check might elucidate whether it would be a good idea to add the building
information into the model.

```{r ppc-group_means}
ppc_stat_grouped(
  y = stan_dat_simple$complaints, 
  yrep = y_rep, 
  group = pest_data$building_id, 
  stat = 'mean',
  binwidth = 0.2
)
```

We're getting plausible predictions for most building means but some are
estimated better than others and some have larger uncertainties than we might
expect. If we explicitly model the variation across buildings we may be able to
get much better estimates.

## Hierarchical modeling

### Modeling varying intercepts for each building

Let's add a hierarchical intercept parameter, $\alpha_b$ at the building level
to our model.

$$
\text{complaints}_{b,t} \sim \text{Neg-Binomial}(\lambda_{b,t}, \phi) \\
\lambda_{b,t}  = \exp{(\eta_{b,t})} \\
\eta_{b,t} = \mu_b + \beta \, {\rm traps}_{b,t} + \beta_{\rm super}\, {\rm super}_b + \text{log_sq_foot}_b \\
\mu_b \sim \text{Normal}(\alpha, \sigma_{\mu})
$$

In our Stan model, $\mu_b$ is the $b$-th element of the vector
$\texttt{mu}$ which has one element per building.

One of our predictors varies only by building, so we can rewrite the above model
more efficiently like so:

$$
\eta_{b,t} = \mu_b + \beta \, {\rm traps}_{b,t} + \text{log_sq_foot}_b\\
\mu_b \sim \text{Normal}(\alpha +  \beta_{\text{super}} \, \text{super}_b , \sigma_{\mu})
$$


We have more information at the building level as well, like the average age of
the residents, the average age of the buildings, and the average per-apartment
monthly rent so we can add that data into a matrix called
`building_data`, which will have one row per building and four columns:

* `live_in_super`
* `age_of_building`
* `average_tentant_age`
* `monthly_average_rent`

We'll write the Stan model like:

$$
\eta_{b,t} = \mu_b + \beta \, {\rm traps} + \text{log_sq_foot}\\
\mu \sim \text{Normal}(\alpha + \texttt{building_data} \, \zeta, \,\sigma_{\mu})
$$

### Prepare building data for hierarchical

We'll need to do some more data prep before we can fit our models. Firstly to
use the building variable in Stan we will need to transform it from a factor
variable to an integer variable.

```{r prep-data}
N_months <- length(unique(pest_data$date))

# Add some IDs for building and month
pest_data <- pest_data %>%
  mutate(
    building_fac = factor(building_id, levels = unique(building_id)),
    building_idx = as.integer(building_fac),
    ids = rep(1:N_months, N_buildings),
    mo_idx = lubridate::month(date)
  )

# Center and rescale the building specific data
building_data <- pest_data %>%
    select(
      building_idx,
      live_in_super,
      age_of_building,
      total_sq_foot,
      average_tenant_age,
      monthly_average_rent
    ) %>%
    unique() %>%
    arrange(building_idx) %>%
    select(-building_idx) %>%
    scale(scale=FALSE) %>%
    as.data.frame() %>%
    mutate( # scale by constants
      age_of_building = age_of_building / 10,
      total_sq_foot = total_sq_foot / 10000,
      average_tenant_age = average_tenant_age / 10,
      monthly_average_rent = monthly_average_rent / 1000
    ) %>%
    as.matrix()

# Make data list for Stan
stan_dat_hier <-
  with(pest_data,
        list(complaints = complaints,
             traps = traps,
             N = length(traps),
             J = N_buildings,
             M = N_months,
             log_sq_foot = log(pest_data$total_sq_foot/1e4),
             building_data = building_data[,-3],
             mo_idx = as.integer(as.factor(date)),
             K = 4,
             building_idx = building_idx
             )
        )
```

### Compile and fit the hierarchical model

Let's compile the model.

```{r comp-NB-hier, cache=TRUE, results="hide", message=FALSE}
comp_model_NB_hier <- stan_model('stan_programs/hier_NB_regression.stan')
```

Fit the model to data.

```{r run-NB-hier}
fitted_model_NB_hier <-
  sampling(
  comp_model_NB_hier,
  data = stan_dat_hier,
  chains = 4,
  cores = 4,
  iter = 4000
  )
```

### Diagnostics

We get a bunch of warnings from Stan about divergent transitions, which is
an indication that there may be regions of the posterior that have not been
explored by the Markov chains.

Divergences are discussed in more detail in the course slides as well as
the **bayesplot** (MCMC diagnostics vignette)[http://mc-stan.org/bayesplot/articles/visual-mcmc-diagnostics.html]
and [_A Conceptual Introduction to Hamiltonian Monte Carlo_](https://arxiv.org/abs/1701.02434).

In this
example we will see that we have divergent transitions because we need to
reparameterize our model - i.e., we will retain the overall structure of the
model, but transform some of the parameters so that it is easier for Stan to
sample from the parameter space. Before we go through exactly how to do this
reparameterization, we will first go through what indicates that this is
something that reparameterization will resolve. We will go through:

1. Examining the fitted parameter values, including the effective sample size
2. Traceplots and scatterplots that reveal particular patterns in locations of the divergences.

First let's extract the fits from the model.

```{r}
samps_hier_NB <- rstan::extract(fitted_model_NB_hier)
```

Then we print the fits for the parameters that are of most interest.

```{r print-NB-hier}
print(fitted_model_NB_hier, pars = c('sigma_mu','beta','alpha','phi','mu'))
```
You can see that the effective samples are quite low for many of the parameters
relative to the total number of samples. This alone isn't indicative of the need
to reparameterize, but it indicates that we should look further at the trace
plots and pairs plots. First let's look at the traceplots to see if the divergent
transitions form a pattern.

```{r}
# use as.array to keep the markov chains separate for trace plots
mcmc_trace(
  as.array(fitted_model_NB_hier,pars = 'sigma_mu'),
  np = nuts_params(fitted_model_NB_hier),
  window = c(500,1000)
)
```

Looks as if the divergent parameters, the little red bars underneath the traceplots
correspond to samples where the sampler gets stuck at one parameter value for $\sigma_\mu$.

```{r}
# assign to object so we can compare to another plot later
scatter_with_divs <- mcmc_scatter(
  as.array(fitted_model_NB_hier),
  pars = c("mu[4]", 'sigma_mu'),
  transform = list('sigma_mu' = "log"),
  np = nuts_params(fitted_model_NB_hier)
)
scatter_with_divs
```

What we have here is a cloud-like shape, with most of the divergences clustering
towards the bottom. 

Another way to look at the divergences is via a parallel coordinates plot:

```{r}
parcoord_with_divs <- mcmc_parcoord(
  as.array(fitted_model_NB_hier, pars = c("sigma_mu", "mu")),
  np = nuts_params(fitted_model_NB_hier)
)
parcoord_with_divs
```

Again, we see evidence that our problems concentrate when $\texttt{sigma_mu}$ is small.

### Reparameterize and recheck diagnostics

Instead, we should use the non-centered parameterization for $\mu_b$. We
define a vector of auxiliary variables in the parameters block,
$\texttt{mu_raw}$ that is given a $\text{Normal}(0, 1)$ prior in the model
block. We then make $\texttt{mu}$ a transformed parameter:
We can reparameterize the random intercept $\mu_b$, which is distributed:

$$
\mu_b \sim \text{Normal}(\alpha + \texttt{building_data} \, \zeta, \sigma_{\mu})
$$



```
transformed parameters {
  vector[J] mu;
  mu = alpha + building_data * zeta + sigma_mu * mu_raw;
}
```

This gives $\texttt{mu}$ a
$\text{Normal}(\alpha + \texttt{building_data}\, \zeta, \sigma_\mu)$
distribution, but it decouples the dependence of the density of each element of
$\texttt{mu}$ from $\texttt{sigma_mu}$ ($\sigma_\mu$).
hier_NB_regression_ncp.stan uses the non-centered parameterization for
$\texttt{mu}$. We will examine the effective sample size of the fitted model to
see whether we've fixed the problem with our reparameterization.

Compile the model.

```{r comp-NB-hier-ncp, cache=TRUE}
comp_model_NB_hier_ncp <- stan_model('stan_programs/hier_NB_regression_ncp.stan')
```

Fit the model to the data.

```{r run-NB-hier-ncp}
fitted_model_NB_hier_ncp <- sampling(comp_model_NB_hier_ncp, data = stan_dat_hier, chains = 4, cores = 4)
```

Examining the fit of the new model

```{r n-eff-NB-hier-ncp-check}
print(fitted_model_NB_hier_ncp, pars = c('sigma_mu','beta','alpha','phi','mu'))
```

This has improved the effective sample sizes of $\texttt{mu}$. We extract
the parameters to run our usual posterior predictive checks.

```{r}
scatter_no_divs <- mcmc_scatter(
  as.array(fitted_model_NB_hier_ncp),
  pars = c("mu[4]", 'sigma_mu'),
  transform = list('sigma_mu' = "log"),
  np = nuts_params(fitted_model_NB_hier_ncp)
)
bayesplot_grid(scatter_with_divs, scatter_no_divs,
               grid_args = list(ncol = 2), ylim = c(-11, 1))
```

```{r}
parcoord_no_divs <- mcmc_parcoord(
  as.array(fitted_model_NB_hier_ncp, pars = c("sigma_mu", "mu")),
  np = nuts_params(fitted_model_NB_hier_ncp)
)
bayesplot_grid(parcoord_with_divs, parcoord_no_divs,
               ylim = c(-3, 3))
```

```{r samps-full-hier}
samps_NB_hier_ncp <- rstan::extract(fitted_model_NB_hier_ncp, pars = c('y_rep','inv_phi'))
```

The marginal plot, again.

```{r ppc-full-hier}
y_rep <- as.matrix(fitted_model_NB_hier_ncp, pars = "y_rep")
ppc_dens_overlay(stan_dat_hier$complaints, y_rep[1:200,])
```

This looks quite nice. If we've captured the building-level means well, then the
posterior distribution of means by building should match well with the observed
means of the quantity of building complaints by month.

```{r ppc-group_means-hier}
ppc_stat_grouped(
  y = stan_dat_hier$complaints,
  yrep = y_rep,
  group = pest_data$building_id,
  stat = 'mean',
  binwidth = 0.5
)
```

We weren't doing terribly with the building-specific means before, but now they
are all estimated a bit better by the model. The model is also able to do a
decent job estimating the probability of zero complaints:

```{r}
prop_zero <- function(x) mean(x == 0)
ppc_stat(
  y = stan_dat_hier$complaints,
  yrep = y_rep,
  stat = prop_zero,
  binwidth = 0.025
)

# plot separately for each building
ppc_stat_grouped(
  y = stan_dat_hier$complaints,
  yrep = y_rep,
  group = pest_data$building_id,
  stat = prop_zero,
  binwidth = 0.025
)
```

Predictions by number of traps:

```{r}
ppc_intervals(
  y = stan_dat_hier$complaints,
  yrep = y_rep,
  x = stan_dat_hier$traps
) +
  labs(x = "Number of traps", y = "Number of complaints")
```

Standardized residuals:
```{r}
mean_y_rep <- colMeans(y_rep)
mean_inv_phi <- mean(as.matrix(fitted_model_NB_hier_ncp, pars = "inv_phi"))
std_resid <- (stan_dat_hier$complaints - mean_y_rep) / sqrt(mean_y_rep + mean_y_rep^2*mean_inv_phi)
qplot(mean_y_rep, std_resid) + hline_at(2) + hline_at(-2)
```


### Varying intercepts _and_ varying slopes

We've gotten some new data that extends the number of time points for which we
have observations for each building. This will let us explore how to expand the
model a bit more with varying _slopes_ in addition to the varying intercepts
and also, later, also model temporal variation.

```{r}
stan_dat_hier <- readRDS('data/pest_data_longer_stan_dat.RDS')
```

Perhaps if the levels of complaints differ by building, the coefficient for the
effect of traps on building does too. We can add this to our model and observe
the fit.

$$
\text{complaints}_{b,t} \sim \text{Neg-Binomial}(\lambda_{b,t}, \phi)  \\
\lambda_{b,t} = \exp{(\eta_{b,t})}\\
\eta_{b,t} = \mu_b + \kappa_b \, \texttt{traps}_{b,t} + \text{log_sq_foot}_b \\
\mu_b \sim \text{Normal}(\alpha + \texttt{building_data} \, \zeta, \sigma_{\mu}) \\
\kappa_b \sim \text{Normal}(\beta + \texttt{building_data} \, \gamma, \sigma_{\kappa})
$$

Let's compile the model.

```{r comp-NB-hier-slopes, cache=TRUE, results="hide", message=FALSE}
comp_model_NB_hier_slopes <- stan_model('stan_programs/hier_NB_regression_ncp_slopes_mod.stan')
```

Fit the model to data and extract the posterior draws needed for
our posterior predictive checks.

```{r run-NB-hier-slopes}
fitted_model_NB_hier_slopes <-
  sampling(
    comp_model_NB_hier_slopes,
    data = stan_dat_hier,
    chains = 4, cores = 4,
    control = list(adapt_delta = 0.95)
  )
```

To see if the model infers building-to-building differences in, we can plot a
histogram of our marginal posterior distribution for `sigma_kappa`.

```{r}
mcmc_hist(
  as.matrix(fitted_model_NB_hier_slopes, pars = "sigma_kappa"),
  binwidth = 0.005
)
```

```{r}
print(fitted_model_NB_hier_slopes, pars = c('kappa','beta','alpha','phi','sigma_mu','sigma_kappa','mu'))
```

```{r}
mcmc_hist(
  as.matrix(fitted_model_NB_hier_slopes, pars = "beta"),
  binwidth = 0.005
)
```

While the model can't specifically rule out zero from the posterior, it does have mass at small
non-zero numbers, so we should leave in the hierarchy over $\texttt{kappa}$. Plotting the marginal
data density again, we can see the model still looks well calibrated.

```{r ppc-full-hier-slopes}
y_rep <- as.matrix(fitted_model_NB_hier_slopes, pars = "y_rep")
ppc_dens_overlay(
  y = stan_dat_hier$complaints,
  yrep = y_rep[1:200,]
)
```


## Time varying effects and structured priors

We haven't looked at how cockroach complaints change over time. Let's look
at whether there's any pattern over time.

```{r ppc-group_max-hier-slopes-mean-by-mo}
select_vec <- which(stan_dat_hier$mo_idx %in% 1:12)
ppc_stat_grouped(
  y = stan_dat_hier$complaints[select_vec],
  yrep = y_rep[,select_vec],
  group = stan_dat_hier$mo_idx[select_vec],
  stat = 'mean'
) + xlim(0, 11)
```

We might augment our model with a log-additive monthly effect, $\texttt{mo}_t$.

$$
\eta_{b,t} = \mu_b + \kappa_b \, \texttt{traps}_{b,t} + \texttt{mo}_t + \text{log_sq_foot}_b
$$

We have complete freedom over how to specify the prior for $\texttt{mo}_t$. There are
several competing factors for how the number of complaints might change over time. It makes
sense that there might be more roaches in the environment during the summer, but we might
also expect that there is more roach control in the summer as well. Given that we're
modeling complaints, maybe after the first sighting of roaches in a building, residents are
more vigilant, and thus complaints of roaches would increase.

This can be a motivation for using an autoregressive prior for our monthly effects. The model
is as follows:

$$
\texttt{mo}_t \sim \text{Normal}(\rho \, \texttt{mo}_{t-1}, \sigma_\texttt{mo}) \\
\equiv \\
\texttt{mo}_t = \rho \, \texttt{mo}_{t-1} +\epsilon_t , \quad \epsilon_t \sim \text{Normal}(0, \sigma_\texttt{mo}) \\
\quad \rho \in [-1,1]
$$

This equation says that the monthly effect in month $t$ is directly related to
the last month's monthly effect. Given the description of the process above,
it seems like there could be either positive or negative associations between
the months, but there should be a bit more weight placed on positive
$\rho$s, so we'll put an informative prior that pushes the parameter $\rho$ towards
0.5.

Before we write our prior, however, we have a problem: Stan doesn't implement any densities
that have support on $[-1,1]$. We can use variable transformation of a raw variable
defined on $[0,1]$ to to give us a density on $[-1,1]$. Specifically,

$$
\rho_{\text{raw}} \in [0, 1] \\
\rho = 2 \times \rho_{\text{raw}} - 1
$$

Then we can put a beta prior on $\rho_\text{raw}$ to push our estimate towards 0.5.

One further wrinkle is that we have a prior for $\texttt{mo}_t$ that depends on
$\texttt{mo}_{t-1}$. That is, we are working with the _conditional_ distribution
of $\texttt{mo}_t$ given $\texttt{mo}_{t-1}$.
But what should we do about the prior for $\texttt{mo}_1$, for which we don't
have a previous time period in the data?

We need to work out the _marginal_ distribution of the first observation.
Thankfully we can use the fact that AR models are stationary, so
$\text{Var}(\texttt{mo}_t) = \text{Var}(\texttt{mo}_{t-1})$ and
$\mathbb{E}(\texttt{mo}_t) = \mathbb{E}(\texttt{mo}_{t-1})$ for all $t$.
Therefore the marginal distribution of $\texttt{mo}_1$ is the same as the
marginal distribution of any $\texttt{mo}_t$.

First we derive the marginal variance of $\texttt{mo}_{t}$.

$$
\text{Var}(\texttt{mo}_t) = \text{Var}(\rho \texttt{mo}_{t-1} + \epsilon_t)  \\
\text{Var}(\texttt{mo}_t) = \text{Var}(\rho \texttt{mo}_{t-1}) + \text{Var}(\epsilon_t)
$$
where the second line holds by independence of $\epsilon_t$ and $\epsilon_{t-1})$.
Then, using the fact that $Var(cX) = c^2Var(X)$ for a constant $c$ and the fact
that, by stationarity, $\textrm{Var}(\texttt{mo}_{t-1}) = \textrm{Var}(\texttt{mo}_{t})$,
we then obtain:

$$
\text{Var}(\texttt{mo}_t)= \rho^2 \text{Var}( \texttt{mo}_{t})  + \sigma_\texttt{mo}^2 \\
\text{Var}(\texttt{mo}_t) = \frac{\sigma_\texttt{mo}^2}{1 - \rho^2}
$$

For the mean of $\texttt{mo}_t$ things are a bit simpler:

$$
\mathbb{E}(\texttt{mo}_t) = \mathbb{E}(\rho \, \texttt{mo}_{t-1} + \epsilon_t) \\
\mathbb{E}(\texttt{mo}_t) = \mathbb{E}(\rho \, \texttt{mo}_{t-1}) + \mathbb{E}(\epsilon_t) \\
$$

Since $\mathbb{E}(\epsilon_t) = 0$ by assumption we have

$$
\mathbb{E}(\texttt{mo}_t) = \mathbb{E}(\rho \, \texttt{mo}_{t-1})  + 0\\
\mathbb{E}(\texttt{mo}_t) = \rho \, \mathbb{E}(\texttt{mo}_{t}) \\
\mathbb{E}(\texttt{mo}_t) - \rho \mathbb{E}(\texttt{mo}_t) = 0  \\
\mathbb{E}(\texttt{mo}_t) = 0/(1 - \rho)
$$

which for $\rho \neq 1$ yields $\mathbb{E}(\texttt{mo}_{t}) = 0$.

We now have the marginal distribution for $\texttt{mo}_{t}$, which, in our case,
we will use for $\texttt{mo}_1$. The full AR(1) specification is then:

$$
\texttt{mo}_1 \sim \text{Normal}\left(0, \frac{\sigma_\texttt{mo}}{\sqrt{1 - \rho^2}}\right) \\
\texttt{mo}_t \sim \text{Normal}\left(\rho \, \texttt{mo}_{t-1}, \sigma_\texttt{mo}\right) \forall t > 1
$$

```{r comp-NB-hier-mos, cache=TRUE, results="hide", message=FALSE}
comp_model_NB_hier_mos <- stan_model('stan_programs/hier_NB_regression_ncp_slopes_mod_mos.stan')
```

```{r run-NB-hier-slopes-mos}
fitted_model_NB_hier_mos <- sampling(comp_model_NB_hier_mos, data = stan_dat_hier, chains = 4, cores = 4, control = list(adapt_delta = 0.9))
```

In the interest of brevity, we won't go on expanding the model, though we certainly could.
What other information would help us understand the data generating process better? What
other aspects of the data generating process might we want to capture that we're not
capturing now?

As usual, we run through our posterior predictive checks.

```{r ppc-full-hier-mos}
y_rep <- as.matrix(fitted_model_NB_hier_mos, pars = "y_rep")
ppc_dens_overlay(
  y = stan_dat_hier$complaints,
  yrep = y_rep[1:200,]
)
```

```{r}
select_vec <- which(stan_dat_hier$mo_idx %in% 1:12)
ppc_stat_grouped(
  y = stan_dat_hier$complaints[select_vec],
  yrep = y_rep[,select_vec],
  group = stan_dat_hier$mo_idx[select_vec],
  stat = 'mean'
)
```

As we can see, our monthly random intercept has captured a monthly pattern
across all the buildings. We can also compare the prior and posterior for the
autoregressive parameter to see how much we've learned. Here are two different
ways of comparing the prior and posterior visually:

```{r}
# 1) compare draws from prior and draws from posterior
rho_draws <- cbind(
  2 * rbeta(4000, 10, 5) - 1, # draw from prior
  as.matrix(fitted_model_NB_hier_mos, pars = "rho")
)
colnames(rho_draws) <- c("prior", "posterior")
mcmc_hist(rho_draws, freq = FALSE, binwidth = 0.025,
          facet_args = list(nrow = 2)) + xlim(-1, 1)


# 2) overlay prior density curve on posterior draws
gen_rho_prior <- function(x) {
  alpha <- 10; beta <- 5
  a <- -1; c <- 1
  lp <- (alpha - 1) * log(x - a) +
        (beta - 1) * log(c - x) -
        (alpha + beta - 1) * log(c - a) -
         lbeta(alpha, beta)
  return(exp(lp))
}
mcmc_hist(as.matrix(fitted_model_NB_hier_mos, pars = "rho"),
          freq = FALSE, binwidth = 0.01) +
  overlay_function(fun = gen_rho_prior) +
  xlim(-1,1)
```
```{r}
print(fitted_model_NB_hier_mos, pars = c('rho','sigma_mu','sigma_kappa','gamma'))
```

```{r}
ppc_intervals(
  y = stan_dat_hier$complaints,
  yrep = y_rep,
  x = stan_dat_hier$traps
) +
  labs(x = "Number of traps", y = "Number of complaints")
```

It looks as if our model finally generates a reasonable posterior predictive
distribution for all numbers of traps, and appropriately captures the tails of the
data generating process.

# Model comparisons

We use now the LOOIC (leave-one-out information criteria) and the WAIC (Watanabe-Akaike information criteria) contained in the ```loo``` package to compare models. The model yielding the lowest LOIIC/WAIC is the best in terms of predictive accuracy for out-of-sample predictions.
We use here the extended dataset and re-fit the models.

```{r model_comp, results="hide", message=FALSE}
library(loo)

# data (longer dataset)
pest_data <- readRDS('data/pest_data_longer_stan_dat.RDS')
str(pest_data)

## ----describe-data-------------------------------------------------------
N_buildings <- length(unique(pest_data$building_idx))
N_buildings 

## Data acquisition

stan_dat_simple <- list(
  N = length(pest_data$complaints), 
  complaints = pest_data$complaints,
  traps = pest_data$traps
)


## Models fit

# simple poisson
comp_model_P <- 
  stan_model('stan_programs/simple_poisson_regression.stan')
fit_P_real_data <- sampling(comp_model_P, data = stan_dat_simple)

# multiple poisson
stan_dat_simple$log_sq_foot <- pest_data$log_sq_foot
stan_dat_simple$live_in_super <- pest_data$pred_mat[,"live_in_super"]
comp_model_P_mult <- 
  stan_model('stan_programs/multiple_poisson_regression.stan')
fit_model_P_mult_real <- sampling(comp_model_P_mult, data = stan_dat_simple)

# negative binomial
comp_model_NB <- 
  stan_model('stan_programs/multiple_NB_regression.stan')
fitted_model_NB <- sampling(comp_model_NB, data = stan_dat_simple)

# -----------------------
stan_dat_hier <- pest_data
# hier NB regression
comp_model_NB_hier <- 
  stan_model('stan_programs/hier_NB_regression.stan')
fitted_model_NB_hier <-
  sampling(
    comp_model_NB_hier,
    data = stan_dat_hier,
    chains = 4,
    cores = 4,
    iter = 4000
  )

# hier NCP NB regression
comp_model_NB_hier_ncp <- 
  stan_model('stan_programs/hier_NB_regression_ncp.stan')
fitted_model_NB_hier_ncp <- sampling(comp_model_NB_hier_ncp, 
        data = stan_dat_hier, 
        chains = 4, 
        cores = 4,
        iter=4000)

# hier NB slopes
comp_model_NB_hier_slopes <- 
  stan_model('stan_programs/hier_NB_regression_ncp_slopes_mod.stan')
fitted_model_NB_hier_slopes <-
  sampling(
    comp_model_NB_hier_slopes,
    data = stan_dat_hier,
    chains = 4, cores = 4,
    control = list(adapt_delta = 0.95))


## Extract pointwise log-likelihood and
## compute loo and waic

# loo
log_lik_pois <- extract_log_lik(fit_P_real_data)
loo_pois <- loo(log_lik_pois)
waic_pois <- waic(log_lik_pois)

log_lik_mult_pois <- extract_log_lik(fit_model_P_mult_real)
loo_mult_pois <- loo(log_lik_mult_pois)
waic_mult_pois <- waic(log_lik_mult_pois)


log_lik_mult_NB <- extract_log_lik(fitted_model_NB)
loo_mult_NB <- loo(log_lik_mult_NB)
waic_mult_NB <- waic(log_lik_mult_NB)


log_lik_hier_NB <- extract_log_lik(fitted_model_NB_hier)
loo_hier_NB <- loo(log_lik_hier_NB)
waic_hier_NB <- waic(log_lik_hier_NB)

log_lik_hier_NB_ncp <- extract_log_lik(fitted_model_NB_hier_ncp)
loo_hier_NB_ncp <- loo(log_lik_hier_NB_ncp)
waic_hier_NB_ncp <- waic(log_lik_hier_NB_ncp)

log_lik_slopes <- extract_log_lik(fitted_model_NB_hier_slopes)
loo_slopes <- loo(log_lik_slopes)
waic_slopes <- waic(log_lik_slopes)


compare(loo_pois, loo_mult_pois,
        loo_mult_NB, loo_hier_NB,
        loo_hier_NB_ncp, loo_slopes)

looic1<-loo_pois$estimates[3,1]
looic2<-loo_mult_pois$estimates[3,1]
looic3<-loo_mult_NB$estimates[3,1]
looic4<-loo_hier_NB$estimates[3,1]
looic5<-loo_hier_NB_ncp$estimates[3,1]
looic6<-loo_slopes$estimates[3,1]
looics <-c(looic1, looic2, looic3, looic4, looic5,
           looic6)

waic1<-waic_pois$estimates[3,1]
waic2<-waic_mult_pois$estimates[3,1]
waic3<-waic_mult_NB$estimates[3,1]
waic4<-waic_hier_NB$estimates[3,1]
waic5<-waic_hier_NB_ncp$estimates[3,1]
waic6<-waic_slopes$estimates[3,1]
waics <-c(waic1, waic2, waic3, waic4, waic5,
          waic6)

par(xaxt="n", mfrow=c(1,2))
plot(looics, type="b", xlab="", ylab="LOOIC")
par(xaxt="s")
axis(1, c(1:6), c("Pois", "Pois mult", "NB",
                  "Hier NB", "Hier NB ncp", "NB slopes"), las=2)
par(xaxt="n")
plot(waics, type="b", xlab="", ylab="WAIC")
par(xaxt="s")
axis(1, c(1:6), c("Pois", "Pois mult", "NB",
                  "Hier NB", "Hier NB ncp", "NB slopes"), las=2)
```


The model with varying slopes appears to be the best in terms of predictive information criteria.