---
title: "Covid-19: Bayesian hierarchical models"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', warning=FALSE, message=FALSE, fig.asp=0.7, dev='png', global.par = TRUE, dev.args=list(pointsize=10), fig.path = 'figs/')
library(MASS)
```
```{r setup, include=FALSE}
library(knitr)
local({
  hook_plot = knit_hooks$get('plot')
  knit_hooks$set(plot = function(x, options) {
    paste0('\n\n----\n\n', hook_plot(x, options))
  })
})
```

# Problem: Hospitalised patients due to Covid-19

Download the data for the Covid-19 spreading outbreak from the official website of [Protezione Civile](https://github.com/pcm-dpc/COVID-19), by using the following command:

```{r covid, eval = FALSE}
read.csv("https://raw.githubusercontent.com/pcm-dpc/COVID-19/master/dati-regioni/dpc-covid19-ita-regioni.csv")
```

The dataset contains the following variables:

- ```data```: Date of notification
- ```stato```: Country of reference
- ```codice_regione```: Code of the Region (ISTAT 2019)
- ```denominazione_regione```: Name of the Region
- ```lat```: Latitude
- ```long```: Longitude
- ```ricoverati_con_sintomi```: Hospitalised patients with symptoms
- ```terapia_intensiva```: Intensive Care
- ```totale_ospedalizzati```: Total hospitalised patients
- ```isolamento_domiciliare```: Home confinement
- ```totale_positivi```: Total amount of current positive cases (Hospitalised patients + Home confinement)
- ```variazione_totale_positivi```: New amount of current positive cases (totale_positivi current day - totale_positivi previous day)
- ```nuovi_positivi```: New amount of current positive cases (totale_casi current day - totale_casi previous day)
- ```dimessi_guariti```: Recovered
- ```deceduti```: Death
- ```totale_casi```: Total amount of positive cases
- ```tamponi```: Tests performed
- ```casi_testati```: Total number of people tested

Consider your dataset until **1 April 2020**. After performing some explanatory analysis:

1. Build a model for total hospitalized **OR** intensive care units.

2. Evaluate the inclusion of the following covariates:

  + time
  + lockdown measures adopted by the Italian Government (see here for more details [https://en.wikipedia.org/wiki/2020_Italy_coronavirus_lockdown](https://en.wikipedia.org/wiki/2020_Italy_coronavirus_lockdown))
  + number of medical swabs
  + regional membership.

3. Study the temporal trend of your selected response variable. 

4. Check the fit of your final model by using posterior predictive checking tools and comment.

5. [optional] Provide 3/4 days-forward *predictions*.

6. [optional] Compare alternative models in terms of predictive information criteria and comment.










